Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-12 paper.tex

\title{Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
\author[1]{Shing H. Zhan}
\author[2,5]{Katherine Eaton}
\author[3,5]{Anastasia Ignatieva}
\author[1,5]{Benjamin Jeffery}
\author[1,5]{Duncan S. Palmer}
\author[1,5]{Yan Wong}
\author[2]{Carmen L. Murall}
\author[4]{Sarah P. Otto}
\author[1,6]{Jerome Kelleher}
\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
\affil[5]{Preliminary alphabetic ordering}
\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}

@@

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain
% We give the date below in the caption - we could probably remove if either from there
% or from here (I suggest from here)
(downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
%% (covering the interval 8393--21618),  % removed as interval now in the figure title
since other regions of the 
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Early recombinants reported by Jackson et al. (2021)}
\label{sec:jackson_recombs}
% - Early Recombinants from Jackson et al
%     Discuss results from Wide ARG in detail. Go into the details of the
% forward-backward HMM runs for each example. Be much more explicity about the
% single origin for the different groups (maybe give one of Ana's graphs for
% Group A (XA I think?)

By searching for samples combining genomic segments from Alpha (B.1.1.7) and
from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
\citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled
in the United Kingdom up to March 7, 2021. The authors proposed that the
recombinants likely arose from eight independent recombination events
(categorised into groups A to D and four singletons).

% jk: this is the proposal from issue #33

\begin{table} \centering
\begin{tabular}{ll|clll}
\toprule
A Group (XA) & Jackson        &  21,256–21,613 & B.1.177/B.1.1.7 \\
             & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
\midrule
B Group & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
        & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
\midrule
C Group & Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
        & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
\midrule
D Group & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
        & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
\midrule
CAMC-CBA018 & Jackson        &  20,390--21,254 & B.1.177/B.1.1.7 \\
            & \texttt{sc2ts} &  16,177--21,255 & B.1.177/B.1.1.7 \\
\midrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals and parent lineages for a subset of the data reported by \cite{Jackson2021-ik}. [FIXME ADD MORE DETAIL] More detailed results are given in Table~\ref{tab:jackson_supplement}.}
\end{table}

In the wide ARG (sampled to mid-2021), all these sequences are descendants of
recombination nodes, except MILK-103C712, which was removed during
preprocessing (Table 2). Additionally, all except QEUH-1067DEF are HMM-consistent:
that is, the mirrored HMM runs (i.e. forward from 5’ to 3’ end versus backwards from 3’ to 5’ end)
identify the same set of originating mutations, have the same number of breakpoints, and have
equivalent (sequence-identical) parents from forward and backward runs.
Overall, we find excellent agreement between Jackson et al.'s
motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
and (2) the breakpoint intervals (Table 2). In all cases, the two methods
proposed identical or very closely related parent lineages. The general mosaic
genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
are two switches in the HMM copying paths (indicative of two breakpoints). The
breakpoint intervals estimated using the two methods overlap substantially in
all the cases (Table 2). Within each of the groups A to D, the constituent
sequences have identical breakpoint intervals, because they trace back to a
single recombination node in the wide ARG. This result is consistent with the
hypothesis that the sequences of the groups A to D emerged from four separate
recombination events~\cite{Jackson2021-ik}.

>>>>

\title{Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
\author[1]{Shing H. Zhan}
\author[2,5]{Katherine Eaton}
\author[3,5]{Anastasia Ignatieva}
\author[1,5]{Benjamin Jeffery}
\author[1,5]{Duncan S. Palmer}
\author[1,5]{Yan Wong}
\author[2]{Carmen L. Murall}
\author[4]{Sarah P. Otto}
\author[1,6]{Jerome Kelleher}
\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
\affil[5]{Preliminary alphabetic ordering}
\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}

@@

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain
% We give the date below in the caption - we could probably remove if either from there
% or from here (I suggest from here)
(downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
since other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Early recombinants}
\label{sec:jackson_recombs}
% FIXME: this is rubbish, can someone do better?
Although coronaviruses are generally known to recombine [CITATION?]
it was not clear that SARS-CoV-2 recombinants with ongoing transmission
existed early in the pandemic.
\citet{Jackson2021-ik} performed a careful analysis of sequences
circulating in the UK in late 2020 to early 2021, and found
evidence of multiple independent recombination events and onward
transmission.
By searching for samples combining genomic segments from Alpha (B.1.1.7) and
from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
they found 16 recombinant sequences from 8 putative
origins (groups A to D and four singletons).
These findings are closely replicated in both the Wide and Long ARGs.

\begin{table} \centering
\begin{tabular}{lllll}
\toprule
Group        & Sequences & Method & Interval & Parents \\
\midrule
A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
             &     &\texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
\cmidrule{3-5}
B            & 2   & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
             &     &\texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
\cmidrule{3-5}
C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
             &     & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
\cmidrule{3-5}
D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
             &     & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
\bottomrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint
intervals and parent lineages for Groups A-D
reported by \cite{Jackson2021-ik} with the corresponding
recombination events in the Wide ARG.
The second column gives the number of sequences in the group.
Intervals are given in left-inclusive, right-exclusive form.
Details for all 16 sequences are
given in Table~\ref{tab:jackson_supplement}.}
\end{table}

The Wide ARG contains 15 of these 16 recombinant sequences
(sample MILK-103C712 was removed during preprocessing; see
Section~\ref{sec:data_preprocessing}), and the results are summarised
in Tables~\ref{tab:jackson} and \ref{tab:jackson_supplement}.
Table~\ref{tab:jackson} shows the groups of sequences identified by Jackson
et al.\ as likely independent recombination events with onward transmission.
In each case we have a corresponding recombination node in the Wide ARG,
from which all the sequences in the group descend. The parent lineages
and breakpoint intervals agree closely (see
Section~\ref{sec:breakpoint_intervals} for more details on breakpoint
intervals).
For groups B, C and D,
these recombination nodes form clades consisting only of the identified
sequences.
Group A sequences were subsequently given the Pango XA designation
following onward transmission,
and there are 44 XA designated samples in the Wide ARG (including the
4 sequences analysed by Jackson et al.). The Group A recombination
node forms a monophyletic clade of these 44 samples.
Table~\ref{tab:jackson_supplement} shows the details for each of the
16 sequences individually, and showing generally a strong concordance
in mosaic structure and parent lineages
(including CAMC-CB7AB3 which is inferred to have two breakpoints under both
methods).

The Long ARG contains 5 of the sequences: two each from groups A and B
and QEUH-1067DEF. These cluster under three recombination nodes, as expected,
and have identical breakpoint intervals and parental lineage assignments
to those of Wide ARG.
The recombination nodes for Group B and QEUH-1067DEF are ancestral only
to the sequences involved.
The recombination node for group A forms a monophyletic clade of all
5 XA samples present in the Long ARG
(Figure~\ref{fig:pango-simple-origin-graph}A)
